Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2015-04-01 paint-contour-tree.tex
Minor edits
Minor edit

\begin{abstract}
We revisit the classic problem of computing the \emph{contour tree}
of a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
% A level set, or contour, at value $v$ is a connected component of $f^{-1}(v)$. 
% As $v$ is changed, these contours change topology, merge into each other, or split.
The contour tree is a fundamental topological data structure that tracks
the evolution of level sets of $f$ and 
% concise representations of $f$ that track the evolution
% of levels sets, and this contour
% behavior. The vertices of these trees are the critical points of $f$, where the gradient is zero.
% The edges represent changes in the topology of contours.
has numerous applications in data analysis and visualization,

Roughly speaking, all existing algorithms require $\Omega(n\log n)$ time for all instances, since
the first step is a global sort of the values of $f$ (technically, only the critical points must be sorted,
but these can be $\Omega(n)$). 
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first input-sensitive algorithm whose time complexity depends
on the contour tree struture, and avoids the global sort for non-pathological inputs.
Our results match all existing upper bounds, but also
characterize the amount of sorting required to compute the contour tree.
This leads to a linear time algorithm for contour tree computation for many
classes of natural instances. 
%There is a crucial use of a family of binomial heaps to maintain priorities, 
Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree, 
allowing us to complement our algorithmic results with matching lower bounds for contour tree computation.

@@

But this is only a worst-case bound. All previous algorithms begin by sorting (at least) the critical points, and
hence must pay $\Omega(t\log t)$ for all instances. Can one beat this sorting bound for some instances
and characterize what inputs are hard? Our main result gives an affirmative answer. First, we require some technical definitions.

>>>>

\begin{abstract}
We revisit the classic problem of computing the \emph{contour tree}
of a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
% A level set, or contour, at value $v$ is a connected component of $f^{-1}(v)$. 
% As $v$ is changed, these contours change topology, merge into each other, or split.
The contour tree is a fundamental topological data structure that tracks
the evolution of level sets of $f$ and 
% concise representations of $f$ that track the evolution
% of levels sets, and this contour
% behavior. The vertices of these trees are the critical points of $f$, where the gradient is zero.
% The edges represent changes in the topology of contours.
has numerous applications in data analysis and visualization.

Roughly speaking, all existing algorithms require $\Omega(n\log n)$ time for all instances, since
the first step is a global sort of the values of $f$ (technically, only the critical points must be sorted,
but these can be $\Omega(n)$). 
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first input-sensitive algorithm whose time complexity depends
on the contour tree struture, and avoids the global sort for non-pathological inputs.
Our results match all existing upper bounds, but also
characterize the amount of sorting required to compute the contour tree.
This leads to an almost linear (inverse Ackermann overhead) time algorithm for contour tree computation for many
classes of natural instances. 
%There is a crucial use of a family of binomial heaps to maintain priorities, 
Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree, 
allowing us to complement our algorithmic results with matching lower bounds for contour tree computation.

@@

But this is only a worst-case bound. All previous algorithms begin by sorting (at least) the critical points, and
hence must pay $\Omega(t\log t)$ for all instances. Can one beat this sorting bound for some instances
and characterize what inputs are hard? Our main result gives an affirmative answer. 
% First, we require some technical definitions.
